Bayesian Hyper-Heuristics

### 2023-02-05 article/sections/01-introduction.tex
✨ (article): Done, submitting for review

A popular field of focus for studying \acfp{ANN} is the process by which these models are trained. \acp{ANN} are trained by optimisation algorithms known as heuristics. Many different heuristics have been developed and used to train \acp{ANN} \cite{ref:gudise:2003, ref:rakitianskaia:2012, ref:montana:1989}. Each of these heuristics have different search behaviours, characteristics, strengths and weaknesses. Finding the best heuristic to train \acp{ANN} is required in order to yield optimal results. This process is often non-trivial and could be a time-consuming exercise. Selection of the best heuristic to train \acp{ANN}, is often problem specific \cite{ref:allen:1996, ref:drake:2020, ref:pillay:2018}.

A recent suggestion related to the field of meta-learning is to dynamically select and/or adjust the heuristic used throughout the training process. This approach focuses on the hybridisation of learning paradigms. One such form of hybridisation of learning paradigms is that of hybridisation of different \textit{heuristics} as they are applied to some optimisation problem \cite{ref:burke:2013}. These methods are referred to as \acfp{HH} and focus on finding the best heuristic in \textit{heuristic space} to solve a specific problem.

In the general context of optimisation, many different types of \acp{HH} have been implemented and applied to many different problems. However, research on the application of \acp{HH} in the context of \acs{ANN} training is scarce. \citeauthor{ref:nel:2021} \cite{ref:nel:2021} provides the first research in this field, applying a \acs{HH} to \acs{FFNN} training.

This research takes a particular interest in developing a population-based, selection \acs{HH} that makes use of probability theory and Bayesian statistical concepts to guide the heuristic selection process. This research presents a novel \acf{BHH}, a new high-level heuristic that utilises a statistical approach, referred to as \textit{Bayesian analysis}, which combines prior information with new evidence to the parameters of a selection probability distribution. This selection probability distribution is the mechanism by which the \acs{HH} selects appropriate heuristics to train \acfp{FFNN} during the training process.

>>>>

A popular field of focus for studying \acfp{ANN} is the process by which these models are trained. \acp{ANN} are trained by optimisation algorithms known as heuristics. Many different heuristics have been developed and used to train \acp{ANN} \cite{ref:rakitianskaia:2012}. Each of these heuristics have different search behaviours, characteristics, strengths and weaknesses. Finding the best heuristic to train \acp{ANN} is required in order to yield optimal results. This process is often non-trivial and could be a time-consuming exercise. Selection of the best heuristic to train \acp{ANN}, is often problem specific \cite{ref:allen:1996}.

A recent suggestion related to the field of \textit{meta-learning} is to dynamically select and/or adjust the heuristic used throughout the training process. This approach focuses on the hybridisation of learning paradigms. One such form of hybridisation of learning paradigms is that of hybridisation of different \textit{heuristics} as they are applied to some optimisation problem \cite{ref:burke:2013}. These methods are referred to as \acfp{HH} and focus on finding the best heuristic in \textit{heuristic space} to solve a specific problem.

In the general context of optimisation, many different types of \acp{HH} have been implemented and applied to many different problems \cite{ref:burke:2013}. However, research on the application of \acp{HH} in the context of \acs{ANN} training is scarce. \citeauthor{ref:nel:2021} \cite{ref:nel:2021} provides some of the first research in this field, applying a \acs{HH} to \acf{FFNN} training.

This research takes a particular interest in developing a population-based, selection \acs{HH} that makes use of probability theory and Bayesian statistical concepts to guide the heuristic selection process. This research presents a novel \acf{BHH}, a new high-level heuristic that utilises a statistical approach, referred to as \textit{Bayesian analysis}, which combines prior information with new evidence to the parameters of a selection probability distribution. This selection probability distribution is the mechanism by which the \acs{HH} selects appropriate heuristics to train \acp{FFNN} during the training process.

### 2023-02-05 article/sections/03-heuristics.tex
✨ (article): Done, submitting for review

Gradient-based heuristics are sensitive to the problem that they are applied to, with hyper-parameter selection often dominating the research focus \cite{ref:bengio:2000, ref:feurer:2019}. \citeauthor{ref:blum:2003} \cite{ref:blum:2003} mention that since the 1980's, a new kind of approximate algorithm has emerged which tries to combine basic heuristic methods in higher level frameworks aimed at efficiently and effectively exploring a search space. These methods are referred to as \acp{MH}.

The biggest difference between \acp{MH} and gradient-bases heuristics is that
\acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN} during training and is not limited to information about the search space \cite{ref:blum:2003}. Similar to gradient-based heuristics, a number of different meta-heuristics have been used to successfully train \acp{FFNN} \cite{ref:rakitianskaia:2012, ref:vanwyk:2014, ref:espinal:2011, ref:gupta:1999}. This research takes a particular interest in population-based \acp{MH} that have been used to successfully train \acp{FFNN}. These include \acf{PSO} \cite{ref:shi:1998}, \acf{DE} \cite{ref:price:2006} and \acfp{GA} \cite{ref:fraser:1957}.

>>>>

Gradient-based heuristics are sensitive to the problem that they are applied to, with hyper-parameter selection often dominating the research focus \cite{ref:bengio:2000}. \citeauthor{ref:blum:2003} \cite{ref:blum:2003} mention that since the 1980's, a new kind of approximate algorithm has emerged which tries to combine basic heuristic methods in higher level frameworks aimed at efficiently and effectively exploring a search space. These methods are referred to as \acp{MH}.

The biggest difference between \acp{MH} and gradient-bases heuristics is that
\acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN} during training and is not limited to information about the search space \cite{ref:blum:2003}. Similar to gradient-based heuristics, a number of different meta-heuristics have been used to successfully train \acp{FFNN} \cite{ref:rakitianskaia:2012, ref:vanwyk:2014, ref:gupta:1999}. This research takes a particular interest in population-based \acp{MH} that have been used to train \acp{FFNN}. These include \acf{PSO} \cite{ref:shi:1998}, \acf{DE} \cite{ref:price:2006} and \acfp{GA} \cite{ref:fraser:1957}.

### 2023-02-05 article/sections/06-bayesian-hyper-heuristics.tex
✨ (article): Done, submitting for review

This section presents the novel \acs{BHH}. The general concept of the \acs{BHH} can be summarised as follows: The \acs{BHH} implements a high-level heuristic selection mechanism that learns to select the best heuristic from a pool of low-level heuristics, to apply to a population of entities, each implementing a candidate solution to a \acs{FFNN}, with the intent of both optimising the underlying \acs{FFNN} and \acs{FFNN} training process. The \acs{BHH} does so by learning the probability that a given heuristic will perform well at a given stage in the \acs{FFNN} training process. These probabilities are then used as heuristic selection probabilities in the next step of the training process.

@@

The credit assignment strategy is a mechanism that assigns a discrete credit indicator to heuristics that perform well, based on their performance metrics such as loss. The credit assignment strategy implements a component of the ``move acceptance'' process as proposed by \citeauthor{ref:ozcan:2006} \cite{ref:ozcan:2006,ref:ozcan:2008} and addresses the credit assignment problem as discussed by \citeauthor{ref:burke:2010} \cite{ref:burke:2010}. A good credit assignment strategy will correctly allocate credit to the appropriate heuristic-entity combination. This research implements the following credit assignment strategies to choose from: \textit{ibest} (iteration best), \textit{pbest} (personal best), \textit{gbest} (global best), \textit{rbest} (replay window best), and \textit{symmetric}, where credit is assigned to all entity-heuristic combinations, regardless of their performance.

@@

The \acs{BHH} implements a probabilistic, predictive model based on the fundamentals of the Na\"{\i}ve Bayes algorithm. The \acs{BHH} thus distinguishes between the following events: \textbf{$\boldsymbol{H}$}, the event of observing \textit{heuristics}, \textbf{$\boldsymbol{E}$}, the event of observing \textit{entities}, \textbf{$\boldsymbol{C}$}, the event of observing \textit{credit assignments} that indicate that the credit assignment \textit{performance criteria} are met. By Bayes' theorem, the selection mechanism implemented by the \acs{BHH} is given as

@@

It can be said that the \acs{BHH} implements a Gaussian process \cite{ref:gortler:2019}. Since the reselection of heuristics happens at regular intervals, the outcome of a selection in one iteration may influence the outcome of another in the next iteration, making the implementation of the \acs{BHH} a \acs{HMM} \cite{ref:rabiner:1986}.

@@

The following hyper-parameters are implemented by the \acs{BHH}: the \textit{heuristic pool} configures the type of heuristics included in the heuristic pool as gradient-only, meta-heuristic-only or all heuristics, the \textit{population size} specifies the number of entities in the entity pool, the \textit{credit assignment strategy} specifies which credit assignment strategy to use, the \textit{reselection interval} determines the frequency of heuristic reselection, the \textit{replay window size} determines the maximum size of the performance log, the \textit{reanalysis interval} determines the frequency at which Bayesian analysis is applied, the \textit{burn in window size} determines the size of an initial window where experience is simply gathered without reanalysis, and finally, \textit{discounted rewards} and \textit{normalisation} flags toggle scaling modifiers on values assigned by the credit assignment strategies, backwards in the performance log.

>>>>

This section presents the novel \acs{BHH}. The general concept of the \acs{BHH} can be summarised as follows: the \acs{BHH} implements a high-level heuristic selection mechanism that learns to select the best heuristic from a pool of low-level heuristics, to apply to a population of entities, each implementing a candidate solution to a \acs{FFNN}, with the intent of both optimising the underlying \acs{FFNN} and \acs{FFNN} training process. The \acs{BHH} does so by learning the probability that a given heuristic will perform well at a given stage in the \acs{FFNN} training process. These probabilities are then used as heuristic selection probabilities in the next step of the training process.

@@

The credit assignment strategy is a mechanism that assigns a discrete credit indicator to heuristics that perform well, based on their performance metrics such as loss. The credit assignment strategy implements a component of the ``move acceptance'' process as proposed by \citeauthor{ref:ozcan:2006} \cite{ref:ozcan:2006} and addresses the credit assignment problem as discussed by \citeauthor{ref:burke:2010} \cite{ref:burke:2010}. A good credit assignment strategy will correctly allocate credit to the appropriate heuristic-entity combination. This research implements the following credit assignment strategies to choose from: \textit{ibest} (iteration best), \textit{pbest} (personal best), \textit{gbest} (global best), \textit{rbest} (replay window best), and \textit{symmetric}, where credit is assigned to all entity-heuristic combinations, regardless of their performance.

@@

The \acs{BHH} implements a probabilistic predictive model based on the fundamentals of the Na\"{\i}ve Bayes algorithm. The \acs{BHH} thus distinguishes between the following events: \textbf{$\boldsymbol{H}$}, the event of observing \textit{heuristics}, \textbf{$\boldsymbol{E}$}, the event of observing \textit{entities}, and \textbf{$\boldsymbol{C}$}, the event of observing \textit{credit assignments} that indicate that the credit assignment \textit{performance criteria} are met. By Bayes' theorem, the selection mechanism implemented by the \acs{BHH} is given as

@@

It can be said that the \acs{BHH} implements a Gaussian process \cite{ref:gortler:2019}. Since the reselection of heuristics happens at regular intervals, the outcome of a selection in one iteration may influence the outcome of another in the next iteration, making the implementation of the \acs{BHH} a \acf{HMM} \cite{ref:rabiner:1986}.

@@

The following hyper-parameters are implemented by the \acs{BHH}: the \textit{heuristic pool} configures the type of heuristics included in the heuristic pool as, the \textit{population size} specifies the number of entities in the entity pool, the \textit{credit assignment strategy} specifies which credit assignment strategy to use, the \textit{reselection interval} determines the frequency of heuristic reselection, the \textit{replay window size} determines the maximum size of the performance log, the \textit{reanalysis interval} determines the frequency at which Bayesian analysis is applied, the \textit{burn in window size} determines the size of an initial window where experience is simply gathered without reanalysis, and finally, the \textit{discounted rewards} and \textit{normalisation} flags toggle scaling modifiers on values assigned by the credit assignment strategies, backwards in the performance log.

### 2023-02-05 article/sections/08-results.tex
✨ (article): Done, submitting for review

To illustrate the learning process undergone by the \acs{BHH}, further investigation is required. This can be achieved by analysing the outcomes of the change in concentration parameters and the probabilitiy distributions implemented by the \acs{BHH}. Figures \ref{fig:results:case_study:alphas:6} to \ref{fig:results:case_study:p_HgEC:0:6} provide illustrations of the change in concentration parameter, $\boldsymbol{\alpha}$ and the effect is has on the probability distributions of heuristic selection probabilities, denoted $P(\boldsymbol{\theta} \vert \boldsymbol{\alpha})$, the prior heuristic selection probability distribution, denoted $P(\boldsymbol{H} \vert \boldsymbol{\theta})$, and the posterior heuristic selection probability distribution, denoted $P(\boldsymbol{H} \vert \boldsymbol{E}, \boldsymbol{C}; \boldsymbol{\theta}, \boldsymbol{\phi}, \boldsymbol{\psi})$. These illustrations represent an example heuristic, \acs{Adam}, and an example entity, identified by index 6 and 0 respectively in the heuristic and entity pools. The illustrations are obtained from 30 runs of the case study on the behaviour of the \acs{BHH} on the iris dataset, illustrated in log scale. Similar illustrations for the rest of the concentration parameters and probability distributions are left out for brevity as they contain similar illustrations. The following sections provide interpretations of each of these concepts.

@@

The concentration parameter $\boldsymbol{\alpha}$ parameterises the Dirichlet probability distribution, denoted $P(\boldsymbol{\theta} \vert \boldsymbol{\alpha})$. The probability distribution, $P(\boldsymbol{\theta} \vert \boldsymbol{\alpha})$, is used to sample prior \textit{heuristic selection probabilities}.

@@

Both the bhh\_baseline (\textcolor{red}{red}) and the bhh\_replay\_250 (\textcolor{green}{green}) configurations make use of the \textit{ibest} credit assignment strategy. The \textit{ibest} credit assignment strategy allocates credit to the heuristic that yields the best performance for the current iteration/step and thus, towards the end of the training process, any random heuristic can yield the best iteration performance. However, since the bhh\_baseline is configured with a small replay window size of 10, and a reanalysis interval of 10, the concentration parameter $\boldsymbol{\alpha}$ is reset to its default value of 1.0 more often than the bhh\_replay\_250 configuration, resulting in a probability distribution that is broader, and thus, explaining the larger variance of $\boldsymbol{\theta}$ throughout.

@@

Furthermore, it can be observed for small windows, at various steps for multiple runs, that the variance of $\boldsymbol{\theta}$, for the bhh\_baseline configuration and the bhh\_credit\_symmetric configuration, do not yield means that are equal to the expected heuristic selection probabilities of 0.1. This is proof that the \acs{BHH} does not just implement a form of random search, despite having small reanalysis interval and replay window size configurations. This is also true for the bhh\_credit\_symmetric configuration, as the bhh\_credit\_symmetric configuration biases towards heuristics that happen to be sampled, despite not biasing towards good performance.

@@

The main observation to make from these figures is that they are much less volatile and noisy than the figures presented for the distribution of heuristic selection probabilities, $\boldsymbol{\theta}$, presented in the previous section. This is because of the \textit{reselection} interval hyper-parameter. The reselection interval hyper-parameter is implemented as a way to control the frequency by which new heuristics are selected and allocated to each entity. Since the default \textit{reselection} interval is set to 10, the heuristic selection probabilities are only resampled at intervals of 10. This illustration then simply yield rough approximation of the illustration provided in Figure \ref{fig:results:case_study:thetas:6}. As such, the same observations and conclusions that are made in Section \ref{sec:results:case_study:probabilities_of_probabilities} also apply in this section.

Finally, it should be mentioned that at each step, the goal of the \acs{BHH} is to update these prior ``beliefs'' based on newly observed evidence of heuristic performances. Since these prior heuristic selection probabilities change over time, it can be concluded that the change in prior heuristic selection probabilities is a result of the learning mechanism of the \acs{BHH}. Furthermore, the prior heuristic selection probability distribution provides an opportunity to utilise prior knowledge by some expert before training starts, by injecting heuristic selection biases, by means of the initial values for the concentration parameter $\boldsymbol{\alpha}$.

@@

The main observation to make from Figure \ref{fig:results:case_study:p_HgEC:0:6} is that the implemented posterior heuristic selection distribution does not yield normalised probabilities, but rather yield unnormalised \textit{logits}, which are used to parameterise a Categorical probability distribution from which new heuristic selections are sampled. The reasons for the aforementioned is because the probabilistic model is evaluated proportionally as was discussed in Section \ref{sec:bhh}. Furthermpre, the log-sum-exp trick is used in order to maintain numerical stability, yielding \textit{logits} instead of probabilities.

>>>>

To illustrate the learning process undergone by the \acs{BHH}, further investigation is required. This can be achieved by analysing the outcomes of the change in concentration parameters and the probabilitiy distributions implemented by the \acs{BHH}. Figures \ref{fig:results:case_study:alphas:6} to \ref{fig:results:case_study:p_HgEC:0:6} provide illustrations of the change in concentration parameter, $\boldsymbol{\alpha}$, and the effect is has on the probability distributions of heuristic selection probabilities, denoted $P(\boldsymbol{\theta} \vert \boldsymbol{\alpha})$, the prior heuristic selection probability distribution, denoted $P(\boldsymbol{H} \vert \boldsymbol{\theta})$, and the posterior heuristic selection probability distribution, denoted $P(\boldsymbol{H} \vert \boldsymbol{E}, \boldsymbol{C}; \boldsymbol{\theta}, \boldsymbol{\phi}, \boldsymbol{\psi})$. These illustrations represent an example heuristic, \acs{Adam}, and an example entity, identified by indices 6 and 0 respectively in the heuristic and entity pools. The illustrations are obtained from 30 runs of the case study on the behaviour of the \acs{BHH} on the iris dataset, illustrated in log scale. Similar illustrations for the rest of the concentration parameters and probability distributions are left out for brevity as they contain similar illustrations. The following sections provide interpretations of each of these concepts.

@@

The concentration parameter, $\boldsymbol{\alpha}$, parameterises the Dirichlet probability distribution, denoted $P(\boldsymbol{\theta} \vert \boldsymbol{\alpha})$. The probability distribution, $P(\boldsymbol{\theta} \vert \boldsymbol{\alpha})$, is used to sample prior \textit{heuristic selection probabilities}.

@@

Both the bhh\_baseline (\textcolor{red}{red}) and the bhh\_replay\_250 (\textcolor{green}{green}) configurations make use of the \textit{ibest} credit assignment strategy. The \textit{ibest} credit assignment strategy allocates credit to the heuristic that yields the best performance for the current iteration/step and thus, towards the end of the training process, any random heuristic can yield the best iteration performance. However, since the bhh\_baseline is configured with a small replay window size of 10, and a reanalysis interval of 10, the concentration parameter, $\boldsymbol{\alpha}$, is reset to its default value of 1.0 more often than the bhh\_replay\_250 configuration, resulting in a probability distribution that is broader, and thus, explaining the larger variance of $\boldsymbol{\theta}$ throughout.

@@

Furthermore, it can be observed for small windows, at various steps for multiple runs, that the variance of $\boldsymbol{\theta}$, for the bhh\_baseline configuration and the bhh\_credit\_symmetric configuration, does not yield means that are equal to the expected heuristic selection probabilities of 0.1. This is proof that the \acs{BHH} does not just implement a form of random search, despite having small reanalysis interval and replay window size configurations. This is also true for the bhh\_credit\_symmetric configuration, as the bhh\_credit\_symmetric configuration biases towards heuristics that happen to be sampled, despite not biasing towards good performance.

@@

The main observation to make from these figures is that they are much less volatile and noisy than the figures presented for the distribution of heuristic selection probabilities, $\boldsymbol{\theta}$, presented in the previous section. This is because of the \textit{reselection} interval hyper-parameter. The reselection interval hyper-parameter is implemented as a way to control the frequency by which new heuristics are selected and allocated to each entity. Since the default \textit{reselection} interval is set to 10, the heuristic selection probabilities are only resampled at intervals of 10. Figure \ref{fig:results:case_study:p_H:6} then simply yields a rough approximation of the illustration provided in Figure \ref{fig:results:case_study:thetas:6}. As such, the same observations and conclusions that are made in Section \ref{sec:results:case_study:probabilities_of_probabilities} also apply in this section.

Finally, it should be mentioned that at each step, the goal of the \acs{BHH} is to update these prior ``beliefs'' based on newly observed evidence of heuristic performances. Since these prior heuristic selection probabilities change over time, it can be concluded that the change in prior heuristic selection probabilities is a result of the learning mechanism of the \acs{BHH}. Furthermore, the prior heuristic selection probability distribution provides an opportunity to utilise prior knowledge by some expert before training starts, by injecting heuristic selection biases, by means of the initial values for the concentration parameter, $\boldsymbol{\alpha}$.

@@

The main observation to make from Figure \ref{fig:results:case_study:p_HgEC:0:6} is that the implemented posterior heuristic selection distribution does not yield normalised probabilities, but rather yield unnormalised \textit{logits}, which are used to parameterise a Categorical probability distribution from which new heuristic selections are sampled. The reasons for the aforementioned is because the probabilistic model is evaluated proportionally as was discussed in Section \ref{sec:bhh}. Furthermore, the log-sum-exp trick is used in order to maintain numerical stability, yielding \textit{logits} instead of probabilities.
